Endogenous Macrodynamics in Algorithmic Recourse

### 2022-12-15 paper/sections/empirical.rmd
trimmed down

For each dataset and generator, we look at three different types of classifiers, all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier---**Logistic Regression**---implemented as single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier. We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty. They are also the model of choice in @schut2021generating. The actual neural network architectures are kept simple (top half of Table \@ref(tab:architecture)), since we are only marginally concerned with achieving good initial classifier performance.

@@

Ex-ante we expect to see that by construction, Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary as we saw in Figure \@ref(fig:poc). Thus, the choice of a black-box model may have an impact on the paths of the recourse. For generators that use latent space search (REVISE @joshi2019realistic, CLUE @antoran2020getting) or rely on (and have access to) probabilistic models (CLUE @antoran2020getting, Greedy @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples. Of course, this expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty. Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class^[As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated. We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.]. In summary, we expect that the endogenous shifts induced by Wachter outsize those induced by all other generators, since Wachter is the only approach that is not concerned with generating what we have defined as meaningful counterfactuals.

@@

We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification. Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@kaggle2011give], originally consisting of 250,000 instances with 11 numerical attributes. Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable whether credit card clients default on their payment. In its raw form it consists of 23 explanatory variables: 4 categorical features relating to demographic attributes^[These have been omitted from the analysis. See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding. Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019realistic and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.

>>>>

For each dataset and generator, we look at three different types of classifiers, all of them built and trained using `Flux.jl` [@innes2018fashionable]: firstly, a simple linear classifier---**Logistic Regression**---implemented as a single linear layer with sigmoid activation; secondly, a multilayer perceptron (**MLP**); and finally, a **Deep Ensemble** composed of five MLPs following @lakshminarayanan2016simple that serves as our only probabilistic classifier. We have chosen to work with deep ensembles both for their simplicity and effectiveness at modelling predictive uncertainty. They are also the model of choice in @schut2021generating. The network architectures are kept simple (top half of Table \@ref(tab:architecture)), since we are only marginally concerned with achieving good initial classifier performance.

@@

Ex-ante we expect to see that by construction, Wachter will create a new cluster of counterfactual instances in the proximity of the initial decision boundary as we saw in Figure \@ref(fig:poc). Thus, the choice of a black-box model may have an impact on the paths of the recourse. For generators that use latent space search (REVISE @joshi2019realistic, CLUE @antoran2020getting) or rely on (and have access to) probabilistic models (CLUE @antoran2020getting, Greedy @schut2021generating) we expect that counterfactuals will end up in regions of the target domain that are densely populated by training samples. Of course, this expectation hinges on how effective said probabilistic models are at capturing predictive uncertainty. Finally, we expect to see the counterfactuals generated by DiCE to be uniformly spread around the feature space inside the target class^[As we mentioned earlier, the diversity constraint used by DiCE is only effective for when at least two counterfactuals are being generated. We have therefore decided to always generate 5 counterfactuals for each generator and randomly pick one of them.]. In summary, we expect that the endogenous shifts induced by Wachter outsize those of all other generators, since Wachter is not explicitly concerned with generating what we have defined as meaningful counterfactuals.

@@

We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification. Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@kaggle2011give], originally consisting of 250,000 instances with 11 numerical attributes. Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable whether credit card clients default on their payment. In its raw form it consists of 23 explanatory variables: 4 categorical features relating to demographic attributes^[These have been omitted from the analysis. See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding. Both datasets have been used in the literature on AR before (see for example @pawelczyk2021carla, @joshi2019realistic and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.

### 2022-12-15 paper/sections/empirical_2.rmd
trimmed down

- The observed, adverse dynamics are generally speaking of the highest magnitude for the simple linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible. 
- The reduction in model performance appears to be most severe when classes are not perfectly separable or the initial performance of the classifier was weak to begin with.
- With the exception of the Greedy generator, all other generators generally perform somewhat better overall than the baseline (Wachter) as expected.

@@

The findings for the other synthetic datasets are broadly consistent with the observations above. For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts. For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.]. Model shifts and performance deterioration are also quantitatively smaller than in what we can observe in Figure \@ref(fig:syn). The same broadly holds for the Moons data,  For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.

@@

Turning to the real-world data we will go through the findings presented in Figure \@ref(fig:real), where each column corresponds to one of the three data sets. The results shown here are for the deep ensemble, which once again largely resemble those for the MLP. Starting from the top row, perhaps somewhat surprisingly we find no substantial domain shifts. While Latent Space search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.

The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): the estimated PP MMD is statistically significant and large for all datasets. Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data. Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.

>>>>

- The observed dynamics are generally of the highest magnitude for the linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible. 
- The reduction in model performance appears to be most severe when classes are not perfectly separable or the initial model performance was weak to begin with.
- With the exception of the Greedy generator, all other generators generally perform somewhat better overall than the baseline (Wachter) as expected.

@@

The findings for the other synthetic datasets are broadly consistent with the observations above. For the Moons data the same broad patterns emerge, although in this case it is less evident that LS generators induce relatively smaller shifts. For the Circles data, it also appears at first sight that LS search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accurate representation of the data generating process.]. Model shifts and performance deterioration are also quantitatively smaller than in what we can observe in Figure \@ref(fig:syn). The same broadly holds for the Moons data,  For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.

@@

Turning to the real-world data we will go through the findings presented in Figure \@ref(fig:real), where each column corresponds to one of the three data sets. The results shown here are for the deep ensemble, which once again largely resemble those for the MLP. Starting from the top row, perhaps somewhat surprisingly we find no substantial domain shifts. While LS search induces domain shifts that are orders of magnitude higher than for the other generators, they are still small enough to be considered negligible.

The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): the estimated PP MMD is statistically significant and large for all datasets. Here we find no evidence that LS search helps to mitigate model shifts, as we did before for the synthetic data. Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.

### 2022-12-15 paper/sections/limitations.rmd
trimmed down

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. The implementation of such changes is typically described as **gaming** @miller2020strategic, although they need not be driven by adversarial intentions: in Example \@ref(exm:student), student applicants may dutifully focus on acquiring credentials that help them to be admitted to university, but ultimately not to improve their chances of success at completing their degree @barocas2017fairness. Preventing such actions may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. @karimi2021algorithmic.

>>>>

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. The implementation of such changes is typically described as **gaming** @miller2020strategic, although they need not be driven by adversarial intentions: in Example \@ref(exm:student), student applicants may dutifully focus on acquiring credentials that help them to be admitted to university, but ultimately not to improve their chances of success at completing their degree @barocas2017fairness. Preventing such actions may help to avoid the dynamics we have pointed to in this work. Future work would likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. @karimi2021algorithmic.

### 2022-12-15 paper/sections/methodology.rmd
trimmed down

Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: $f$ is just the identity function and the number of counterfactuals $K$ is equal to one. Both REVISE and CLUE search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case, $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019realistic].

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$. To quantify predictive uncertainty, Antoran et al. [@antoran2020getting] rely on entropy estimates for probabilistic models. The greedy approach proposed by Schut et al. @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the cost function $\text{cost}(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. [@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$, we also implicitly minimize predictive entropy (as in CLUE). In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski et al. @dombrowski2021diffeomorphic relies on normalizing flows.].

>>>>

Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is a $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: $f$ is the identity function and the number of counterfactuals $K$ is one. Both REVISE and CLUE search counterfactuals in some latent space $\mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case, $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, that is the function that maps back from the latent space to inputs. Provided the generative model is well-specified, traversing the latent embedding typically yields meaningful counterfactuals, since they are implicitly generated by the (learned) DGP [@joshi2019realistic].

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$. To quantify predictive uncertainty, Antoran et al. [@antoran2020getting] rely on entropy estimates for probabilistic models. The greedy approach proposed by Schut et al. @schut2021generating, which we refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the cost function $\text{cost}(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. [@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$, we also implicitly minimize predictive entropy (as in CLUE). In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski et al. @dombrowski2021diffeomorphic relies on normalizing flows.].

### 2022-12-15 paper/sections/methodology_2.rmd
trimmed down

::: {.proposition #costs name="Costs"}
If so, are any of these dynamics substantial enough to be considered costly to any of the stakeholders involved in real-world automated decision-making processes?
:::

@@

As an alternative to PP MMD, we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement). This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is, the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample. Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries. In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model. A Disagreement Coefficient unequal to zero is indicative of a model shift. The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured. This is one reason for why PP MMD is our our preferred metric.

>>>>

::: {.proposition #costs name="Costs"}
If so, are these dynamics substantial enough to be considered costly to stakeholders involved in real-world automated decision-making processes?
:::

@@

As an alternative to PP MMD, we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement). This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is the probability that two classifiers disagree on the predicted outcome for a randomly chosen sample. Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries. In our context, this metric quantifies the overlap between the initial model (trained before the application of AR) and the updated model. A Disagreement Coefficient unequal to zero is indicative of a model shift. The opposite is not true: even if the Disagreement Coefficient is equal to zero, a model shift may still have occured. This is one reason for why PP MMD is our preferred metric.

### 2022-12-15 paper/sections/mitigation.rmd
trimmed down

for each counterfactual $k$ where $l$ denotes the loss function used to train $M$. This approach, which we shall refer to as **ClaPROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss. It is closely linked to the idea of choosing a higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers. It also makes the private vs. external cost trade-off more explicit and hence manageable.

@@

Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts. Figure \@ref(fig:mitigate-results) presents a subset of the evaluation metrics for our synthetic data with overlapping classes. The top row in Figure \@ref(fig:mitigate-results) indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts. ClaPROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts. As evident from the middle row in Figure \@ref(fig:mitigate-results) model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective. The same overall pattern can be observed for out-of-sample model performance. With respect to the other synthetic datasets, for the Moons dataset, the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.

@@

An interesting finding is also that the proposed strategies have a complementary effect when used in combination with Latent Space generators. In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClaPROAR. In Figure \@ref(fig:mitigate-latent-results) the conventional Latent Space generator with $\gamma=0.5$ serves as our baseline. Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.

```{r mitigate-latent-results, fig.cap="Combinining various mitigation strategies with Latent Space search. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a mesh grid of points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/mitigation_synthetic_latent_results.png")
```

>>>>

for each counterfactual $k$ where $l$ denotes the loss function used to train $M$. This approach, which we refer to as **ClaPROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss. It is closely linked to the idea of choosing a higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers. It also makes the private vs. external cost trade-off more explicit and hence manageable.

@@

Our findings indicate that all three mitigation strategies are at least at par with LS generators with respect to their effectiveness at mitigating domain and model shifts. Figure \@ref(fig:mitigate-results) presents a subset of the evaluation metrics for our synthetic data with overlapping classes. The top row in Figure \@ref(fig:mitigate-results) indicates that while domain shifts are of roughly the same magnitude for both Wachter and LS generators, our proposed strategies effectively mitigate these shifts. ClaPROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts. As evident from the middle row in Figure \@ref(fig:mitigate-results) model shifts can also be reduced: for the deep ensemble LS search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective. The same overall pattern can be observed for out-of-sample model performance. With respect to the other synthetic datasets, for the Moons dataset, the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.

@@

An interesting finding is also that the proposed strategies have a complementary effect when used in combination with LS generators. In experiments we conducted on the synthetic data, the benefits of LS generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClaPROAR. In Figure \@ref(fig:mitigate-latent-results) the conventional LS generator with $\gamma=0.5$ serves as our baseline. Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.

```{r mitigate-latent-results, fig.cap="Combinining various mitigation strategies with LS search. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a mesh grid of points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/mitigation_synthetic_latent_results.png")
```

### 2022-12-15 paper/sections/related.rmd
trimmed down

In the baseline work [@wachter2017counterfactual], the cost function is proxied by some distance metric based on the simple intuition that perturbations of $x$ are costly to the individual. For models that are differentiable and produce smooth predictions, \@ref(eq:solution) can be solved through gradient descent. This summarizes the approach followed in @wachter2017counterfactual which we shall refer to simply as **Wachter**, the name of the first author, in the remainder of this paper.

>>>>

In the baseline work [@wachter2017counterfactual], the cost function is proxied by some distance metric based on the simple intuition that perturbations of $x$ are costly to the individual. For models that are differentiable and produce smooth predictions, \@ref(eq:solution) can be solved through gradient descent. This summarizes the approach followed in @wachter2017counterfactual which we refer to simply as **Wachter**, the name of the first author, in the remainder of this paper.
